---
title: "TrendAnalysis"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries, include = FALSE}
library(HTLNbladderpod) # loading bladder pod data
library(DBI) # loading bladder pod data
library(odbc) # loading bladder pod data
library(tidyverse) # data wrangling
# library(car) # regression diagnostics
library(brms) # bayesian modeling
library(bayesplot) # plotting bayesian
# library(tidybayes) # bayesian tidy stuff
# library(loo) # model selection 
# library(lme4) # model structure
```


The following code creates a model that determines how density class varies with year while accounting for location. 

```{r loading data, include = FALSE}
# getting bloodyhill data ----
importData(instance = 'local',
           path = "C:/Users/kbailey/Documents/HTLN/HTLN Database/",
           name = "MoBlad_BloodyHill_v3.66.accdb",  # "MoBlad_Glades_v1.8.accdb"
           new_env = TRUE)

# merging data
bh_counts <- VIEWS_HTLN_BLDP$tbl_5m_MoBladCount
bh_period <- VIEWS_HTLN_BLDP$tlu_PeriodID
bh_core <- VIEWS_HTLN_BLDP$tlu_5m_Grid_Core
density_class <- VIEWS_HTLN_BLDP$tlu_DensityClasses
bh_accuracy <- VIEWS_HTLN_BLDP$tbl_5m_Accuracy
bh_cedar1 <- VIEWS_HTLN_BLDP$tbl_5m_Cedar

# stem density/basal area per unit area
bh_cedar <- bh_cedar1 |>
  mutate(basal_tree = pi * (dbh/2)^2) |> # cm2
  summarise(count = n(),
            ba_total = sum(basal_tree, na.rm = TRUE),
            .by = c("Period_ID",
                    "Location_ID")) |>
  mutate(ba_m2 = ba_total / 25,
         stemd_m2 = count / 25)


bh_data1 <- bh_counts |>
  dplyr::left_join(bh_period,
                   by = "Period_ID") |>
  dplyr::left_join(bh_core) |>
  dplyr::left_join(bh_accuracy,
                   by = c("Period_ID",
                          "Location_ID")) |>
  dplyr::left_join(density_class,
                   by = "DensityClass") |>
  dplyr::left_join(bh_cedar,
                   by = c("Location_ID",
                          "Period_ID"))

# adding core/non-core column
bh_data <- bh_data1 |>
  # core vs non-core
  mutate(type = if_else(is.na(Core_ID), "non-core", "core")) |>
  # converting classes
  mutate(Date = as.Date(StartDate, format = "%m/%d/%Y"),
         year = year(Date),
         DensityClass_f = factor(DensityClass,
                                 levels = c("0",
                                            "1",
                                            "2",
                                            "3",
                                            "4",
                                            "5",
                                            "6"),
                                 ordered = TRUE),
         min_year = min(year),
         year_cat = year - min_year,
         Location_IDf = factor(Location_ID)) |>
  select(Location_ID,
         Location_IDf,
         DensityClass,
         DensityClass_f,
         ParkCode,
         Date,
         year_cat,
         year,
         CalYr,
         AreaSampled,
         type,
         ba_m2,
         stemd_m2,
         PeriodDescriptor,
         Count,
         DensityClass_Est,
         DensityClass_Actual,
         LowerBound,
         UpperBound,
         Midpoint,
         LowerBound50,
         UpperBound50)

bh_dc_count <- bh_data |>
  dplyr::count(DensityClass,
               year)

bh_dc_count_core <- bh_data |>
  dplyr::filter(type == "core") |>
  dplyr::count(DensityClass,
               year)

bh_dc_avg <- bh_data |>
  dplyr::summarise(mean_dc = mean(DensityClass),
                   .by = year)
bh_dc_avg_type <- bh_data |>
  dplyr::summarise(mean_dc = mean(DensityClass),
                   .by = c(year,
                           type))
```

How do density classes change through time? 
```{r density classes through time}
ggplot() +
  geom_line(data = bh_dc_count,
            aes(year,
                n,
                color = factor(DensityClass))) + 
  labs(x = "Year",
       y = "Count") + 
  scale_color_discrete(name = "Density Class") +
  theme_bw()
```

```{r average density class through time}
ggplot() +
  geom_line(data = bh_dc_avg,
            aes(year,
                mean_dc)) +
  labs(x = "Year",
       y = "Average Density Class") +
  theme_bw()
```

```{r hist density class each year}
ggplot() +
  geom_bar(data = bh_data,
           aes(DensityClass_f)) +
  facet_wrap(~year) +
  labs(x = "Density Class",
       y = "Count") +
  theme_bw()
```

Building the model: 
Density Class ~ Year + (1|Location)

```{r year_model, inclue = FALSE}
model_brms <- readRDS("model_brms.rds")
```

Looking at Posterior Analysis

Model Summary:
```{r year_model_summary}
summary(model_brms)
```
- R_hat
  - Year (Fixed Effects) = 1.00
  - Location (Random Effects) = 1.01
  - Thresholds = 1.01, 1.01, 1.01, 1.01, 1.00, 1.00


Trace Plots: good
```{r year_trace_plots}
plot(model_brms)
```

Effective sample size: mostly good
```{r year_neff_ratio}
summary(neff_ratio(model_brms))
```

Divergent transitions: zero
```{r year_divergent_transitions}
# all samples with divergent transitions
model_brms |>
  tidybayes::tidy_draws() |>
  dplyr::select(ends_with("__")) |>
  filter(divergent__ == TRUE) # no divergent transitions
```

Posterior Predictive Checks: 
- Does the model reproduce the observed Density Classes?

```{r year_ppc}
pp_check(model_brms,
         type = "bars",
         ndraws = 100)
```

Leave one out cross validation
```{r year_loo_val, warning = FALSE}
loo(model_brms)
```

- 99.9% of observations have good Pareto k values 
- p_loo = 675.3 (flexible but not overfitting)

Model Interpretation: 

- Density varies strongly by location (sd(Intercept) = 2.43)
  - Location differs from another by +- 2.43 units in latent density
- There is a 3% increase in the odds of being in a higher density class every additional year.
  - Estimate = 0.03, CrI = [0.03, 0.04]

Effects Plot: 
```{r year_conditional_effects}
conditional_effects(model_brms,
                    categorical = TRUE)
```
- For every year increase, the probability of being in Density Class 0 decreases. 
- For every year increase, the probability of being in Density Class 1 increases (same for 2, 3, and 4, just slightly less)



The following code creates a model that determines how density class varies with increases in cedar while accounting for location. 

Building the model: 
Density Class ~ Year + Basal Area + (1|Location)

```{r model_cedar, inclue = FALSE}
model_cedar <- readRDS("model_cedar.rds")
```

Looking at Posterior Analysis

Model Summary:
```{r cedar_model_summary}
summary(model_cedar)
```
- R_hat
  - ba_m2 = 1.00
  - Year (Fixed Effects) = 1.00
  - Location (Random Effects) = 1.01
  - Thresholds = 1.0, 1.0, 1.0, 1.0, 1.0, 1.0


Trace Plots: good
```{r cedar_trace_plots}
plot(model_cedar)
```

Effective sample size: mostly good
```{r cedar_neff_ratio}
summary(neff_ratio(model_cedar))
```

Divergent transitions: zero
```{r cedar_divergent_transitions}
# all samples with divergent transitions
model_cedar |>
  tidybayes::tidy_draws() |>
  dplyr::select(ends_with("__")) |>
  filter(divergent__ == TRUE) # no divergent transitions
```

Posterior Predictive Checks: 
- Does the model reproduce the observed Density Classes?

```{r cedar_ppc}
pp_check(model_cedar,
         type = "bars",
         ndraws = 100)
```

Leave one out cross validation
```{r cedar_loo_val}
loo(model_cedar)
```

- 87.7% of observations have good Pareto k values 
- p_loo = 633.7 (flexible but not overfitting)

Model Interpretation: 

- Cedar lowers the latent density score by -0.11 (hardly any)

Effects Plot: 
```{r cedar_conditional_effects}
conditional_effects(model_cedar,
                    categorical = TRUE,
                    "ba_m2")
```
- For every increase in cedar basal area per meter2, the probability of being in Density Class 0 increases. 
- For every increase in cedar basal area per meter2, the probability of being in Density Class 1 decrease (same with 2 and kind of 3, 4).
